---
title: "Linking community science data with population dynamics under stochastic process in the context of ecological release"
author: "Orlando Acevedo-Charry, José Miguel Ponciano, Scott K. Robinson, & Miguel A. Acevedo - Natya Hans is supporting HiperGator opperations and will be more involved for publications"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Could eBird data track population dynamics in time-series in the context of ecological release?

Ecological release: Population level niche expansion and/or shift when
antagonistic interspecific interaction is removed or reduced (Herrmann
*et al.*, *TREE* 2021). Hypothetical ecological release could be
explored as a function of species richness. Lower competition in islands
expand niche (*e.g.*, abundance), while more habitats and more species
increase level of specialization in the Mainland (Sherry *et al.*, *Auk*
2020).

Using the Caribbean meta-archipelago, we filtered the no-aquatic birds
reported in eBird, between Julian day 125 and 150 from 2012-2013,
between 6 and 11 hours in the morning. We excluded incomplete list and
retained checklists of only "traveling" or "stationary" protocol, with a
minimum effort of less than 2 km, more than 10 minutes but less than 2
hours (10:120 min) and less than 10 observers. We assumed as counting
samples the mean of counts of each species reported by the observers
during the 25 days filtered per year; if the mean is very constant among
independent observers, we assume a higher probability of detection,
while higher variation of the counts might reflect a difference in
detection probability. These data constitutes the primary time-series of
species for different islands.

To test a model of stochasticity process, we selected a widespread and
easy to detect species in the meta-archipelago, which are distributed
from the Mainland to different islands in the Caribbean: Bananaquit
*Coereba flaveola*. Theory predicts that species in the mainland have
lower density than islands, and that density could increase as species
richness and potentially interspecific competition decreases.

### Filtering eBird data in meta-archipelagos

The code in HiperGator was (Do not run):

```{r eval=FALSE}
########################################
###~ eBird data - Island filtering ~####
########################################

#Packages
library(auk); #eBird data filters
library(tidyverse); #manipulate data
library(dggridR) #Spatiotemporal Subsampling
library(iNEXT) #rarefaction

#~ eBird data ~####
#Data downloaded from eBird, up to August 2023, saved in the folder of the HPG

#Define the range
CaribbeanRange <- c(-91,8,-59,27)
IndoMalayRange <- c(90,-20,180,20)

#select the columns to extract (based on x$col_idx$name)
colsE <- c("observer_id", "sampling_event_identifier",
           "group identifier",
           "common_name", "scientific_name",
           "observation_count",
           "country", "state_code", "locality_id", "latitude", "longitude",
           "protocol_type", "all_species_reported",
           "observation_date",
           "time_observations_started",
           "duration_minutes", "effort_distance_km",
           "number_observers")

#For Caribbean
f_ebdCa <- "ebd_IslandsCa.txt" #Temporal file to save the filtering eBird data (records)
f_sedCa <- "sed_IslandsCa.txt" #Temporal file to save the filtering sampling event data

ebd_filtCarib<-auk_ebd("eBirddata/ebd_relAug-2023.txt",
                       file_sampling = "eBirddata/ebd_sampling_relAug-2023.txt") %>%
  auk_bbox(CaribbeanRange) %>% #W, S, E, N 
  auk_year(c(2012:2023)) %>%
  auk_protocol(c("Traveling", "Stationary")) %>%
  auk_distance(distance = c(0,5)) %>%
  auk_duration(duration = c(0,300))%>%
  auk_complete() %>%
  auk_filter(f_ebdCa, f_sedCa, overwrite=T, keep = colsE)

f_ebdIM <- "eBirddata/ebd_IslandsIM.txt" #Temporal file to save the filtering eBird data (records)
f_sedIM <- "eBirddata/sed_IslandsIM.txt" #Temporal file to save the filtering sampling event data

ebd_filtIndoP<-auk_ebd("eBirddata/ebd_relAug-2023.txt",
                       file_sampling = "eBirddata/ebd_sampling_relAug-2023.txt") %>%
  auk_bbox(IndoMalayRange) %>% #W, S, E, N 
  auk_year(c(2012:2023)) %>%
  auk_protocol(c("Traveling", "Stationary")) %>%
  auk_distance(distance = c(0,5)) %>%
  auk_duration(duration = c(0,300))%>%
  auk_complete() %>%
  auk_filter(f_ebdIM, f_sedIM, overwrite=T, keep = colsE)

#and with read_ebd I apply another filter to do not repeat records from groups
sed_Carib <- read_sampling(f_sedCa)
ebd_Carib <- read_ebd(f_ebdCa) 
sed_IndoM <- read_sampling(f_sedIM)
ebd_IndoM <- read_ebd(f_ebdIM) 

# Function to convert time observation to hours since midnight
time_to_decimal <- function(x) {
  x <- hms(x, quiet = TRUE)
  hour(x) + minute(x) / 60 + second(x) / 3600
}

#Caribbean data

Carib_eff <- ebd_Carib %>%
  mutate(
    # I don't have here count in X to convert to NA
    observation_count = as.integer(observation_count),
    # effort_distance_km to 0 for non-travelling counts
    effort_distance_km = if_else(protocol_type == "Stationary",
                                 0, effort_distance_km),
    # convert time to decimal hours since midnight
    time_observations_started = time_to_decimal(time_observations_started),
    hour_sampling = round(time_observations_started, 0),
    # split date into year, month, week, and day of year
    year = year(observation_date),
    month = month(observation_date),
    week = week(observation_date),
    day_of_year = yday(observation_date)) %>%
  filter(number_observers <= 10,         #Only list with less than 10 observers
         effort_distance_km <= 5,        #be sure of distance effort
         duration_minutes %in% (0:300), #be sure of duration effort
         !is.na(observation_count))      #only records with abundance estimation

Checklists_with_more_than_fourSppC <- Carib_eff %>%
  select(sampling_event_identifier, scientific_name) %>%
  group_by(sampling_event_identifier) %>%
  summarise(N_species = n()) %>%
  filter(N_species >= 4)
#352797 lists

Caribbean_cooc <- Carib_eff[Carib_eff$sampling_event_identifier %in% Checklists_with_more_than_fourSppC$sampling_event_identifier, ] 

Carib_sed <- sed_Carib %>%
  mutate(
    # effort_distance_km to 0 for non-travelling counts
    effort_distance_km = if_else(protocol_type == "Stationary",
                                 0, effort_distance_km),
    # convert time to decimal hours since midnight
    time_observations_started = time_to_decimal(time_observations_started),
    hour_sampling = round(time_observations_started, 0),
    # split date into year, month, week, and day of year
    year = year(observation_date),
    month = month(observation_date),
    week = week(observation_date),
    day_of_year = yday(observation_date)) %>%
  filter(number_observers <= 10,         #Only list with less than 10 observers
         effort_distance_km <= 5,        #be sure of distance effort
         duration_minutes %in% (0:300)) #be sure of duration effort

#And now with the Oriental-Indo_Malayan-Papuan-Melanesian
IndoM_eff <- ebd_IndoM %>%
  mutate(
    # I don't have here count in X to convert to NA
    observation_count = as.integer(observation_count),
    # effort_distance_km to 0 for non-travelling counts
    effort_distance_km = if_else(protocol_type == "Stationary",
                                 0, effort_distance_km),
    # convert time to decimal hours since midnight
    time_observations_started = time_to_decimal(time_observations_started),
    hour_sampling = round(time_observations_started, 0),
    # split date into year, month, week, and day of year
    year = year(observation_date),
    month = month(observation_date),
    week = week(observation_date),
    day_of_year = yday(observation_date)) %>%
  filter(number_observers <= 10,         #Only list with less than 10 observers
         effort_distance_km <= 5,        #be sure of distance effort
         duration_minutes %in% (0:300), #be sure of duration effort
         !is.na(observation_count))      #remove records without abundance estimation  

Checklists_with_more_than_fourSppIP <- IndoM_eff %>%
  select(sampling_event_identifier, scientific_name) %>%
  group_by(sampling_event_identifier) %>%
  summarise(N_species = n()) %>%
  filter(N_species >= 4)

IndoM_eff_cooc <- IndoM_eff[IndoM_eff$sampling_event_identifier %in% Checklists_with_more_than_fourSppIP$sampling_event_identifier, ]


IndoM_sed <- sed_IndoM %>%
  mutate(
    # effort_distance_km to 0 for non-travelling counts
    effort_distance_km = if_else(protocol_type == "Stationary",
                                 0, effort_distance_km),
    # convert time to decimal hours since midnight
    time_observations_started = time_to_decimal(time_observations_started),
    hour_sampling = round(time_observations_started, 0),
    # split date into year, month, week, and day of year
    year = year(observation_date),
    month = month(observation_date),
    week = week(observation_date),
    day_of_year = yday(observation_date)) %>%
  filter(number_observers <= 10,         #Only list with less than 10 observers
         effort_distance_km <= 5,        #be sure of distance effort
         duration_minutes %in% (0:300)) #be sure of duration effort

#Only those land birds ####
#AVONET filtered by No aquatics
filterAvonet <- read.csv("filterAVONET/AVONET_filter_NoAquatics.csv")

#to combine the data, we have to adjust the names
names(filterAvonet)[2] <- "scientific_name"

#filter in eBird data those species identify in AVONET as land birds
ebird.carib = Caribbean_cooc[Caribbean_cooc$scientific_name %in% filterAvonet$scientific_name, ]
ebird.indop = IndoM_eff_cooc[IndoM_eff_cooc$scientific_name %in% filterAvonet$scientific_name, ]

##Spatial Grids ####
set.seed(123)

# Spatiotemporal Subsampling in diameters of ~1 km (area of ~100 km^2)
dggs_island <- dgconstruct(spacing = 11) #spacing 1 correspond to Characteristic Length Scale, or diameter of spherical cell

#add a new variable that identify cell
Caribbean_SS <- ebird.carib %>%
  mutate(cell = dgGEO_to_SEQNUM(dggs_island, #id for cells
                                longitude, latitude)$seqnum) %>%
  group_by(cell, scientific_name) %>%
  mutate(mu_count = mean(observation_count))

IndoPacific_SS <- ebird.indop %>%
  filter(observation_count < 7000) %>% #there is an outlying count for _Eurylaimus ochromalus_ https://ebird.org/checklist/S112199250
  mutate(cell = dgGEO_to_SEQNUM(dggs_island, #id for cells
                                longitude, latitude)$seqnum) %>%
  group_by(cell, scientific_name) %>%
  mutate(mu_count = mean(observation_count))

#Get the number of species in each cell
RichnessCellCaribbean   <- Caribbean_SS %>% 
  group_by(cell, scientific_name) %>%
  summarise(count=n()) %>% 
  group_by(cell) %>% 
  summarise(SpRichness=n()) %>%
  filter(SpRichness >= 4)

#Exclude cells in SS with less than 4 species
Caribbean_SS = Caribbean_SS[Caribbean_SS$cell %in% RichnessCellCaribbean$cell, ]

#Community assemblages
Caribbean_Cell_Assembly <- Caribbean_SS %>%
  select(cell, scientific_name, mu_count) %>%
  summarise(mu_count = mean(mu_count)) %>%
  pivot_wider(names_from = cell, values_from = mu_count, values_fill = 0)

#Get the grid cell boundaries for cells which had quakes
gridCaribbean <- dgcellstogrid(dggs_island,RichnessCellCaribbean$cell)

#Update the grid cells' properties to include the number of lists in each cell
gridCaribbean <- merge(gridCaribbean,RichnessCellCaribbean,by.x="seqnum",by.y="cell")

# Handle cells that cross 180 degrees
wrapped_gridCaribbean = st_wrap_dateline(gridCaribbean, 
                                         options = c("WRAPDATELINE=YES","DATELINEOFFSET=180"), quiet = TRUE)

#Backup
saveRDS(wrapped_gridCaribbean, "Completeness_data/wrapped_gridCaribbean.rds")
saveRDS(Caribbean_SS, "Completeness_data/Caribbean_SS_140324.rds")
saveRDS(Caribbean_Cell_Assembly, "Completeness_data/Caribbean_Cell_Assembly.rds")

#Get the number of records in each cell
RichnessCellIndoPacific   <- IndoPacific_SS %>% 
  group_by(cell, scientific_name) %>%
  summarise(count=n()) %>% 
  group_by(cell) %>% 
  summarise(SpRichness=n()) %>%
  filter(SpRichness >= 4)

#Exclude cells in SS with less than 4 species
IndoPacific_SS = IndoPacific_SS[IndoPacific_SS$cell %in% RichnessCellIndoPacific$cell, ]

#Community assemblages
IndoPacific_Cell_Assembly <- IndoPacific_SS %>%
  select(cell, scientific_name, mu_count) %>%
  summarise(mu_count = mean(mu_count)) %>%
  pivot_wider(names_from = cell, values_from = mu_count, values_fill = 0)

#Get the grid cell boundaries for cells which had quakes
gridIndoPacific <- dgcellstogrid(dggs_island,RichnessCellIndoPacific$cell)

#Update the grid cells' properties to include the number of lists in each cell
gridIndoPacific <- merge(gridIndoPacific,RichnessCellIndoPacific,by.x="seqnum",by.y="cell")

# Handle cells that cross 180 degrees (and some weird in the map)
wrapped_gridIndoPacific = st_wrap_dateline(gridIndoPacific, 
                                           options = c("WRAPDATELINE=YES","DATELINEOFFSET=180"), quiet = TRUE)

#Backup
saveRDS(wrapped_gridIndoPacific, "Completeness_data/wrapped_gridIndoPacific.rds")
saveRDS(IndoPacific_SS, "Completeness_data/IndoPacific_SS_140324.rds")
saveRDS(IndoPacific_Cell_Assembly, "Completeness_data/IndoPacific_Cell_Assembly.rds")
 
#End of this code

```

### Estimate sample-coverage completeness per sampling unit

Then, we estimated the completeness by sampling units (Cell id in a grid
of \~$100~ km^2$). This was also conducted in the HiperGator (Do not
run):

```{r eval=FALSE}
### Assess completeness of the sampling units

#* Orlando Acevedo-Charry
#* Updated on March 22nd, 2024


#Packages

library(tidyverse)
library(iNEXT) #rarefaction

#Load the data from Code_Islands_eBird_filtering.R

Caribbean_SS <- readRDS("Completeness_data/Caribbean_SS_140324.rds")
IndoPacific_SS <- readRDS("Completeness_data/IndoPacific_SS_140324.rds")

Caribbean_Cell_Assembly <- readRDS("Completeness_data/Caribbean_Cell_Assembly.rds")
IndoPacific_Cell_Assembly <- readRDS("Completeness_data/IndoPacific_Cell_Assembly.rds")

wrapped_gridCaribbean <- readRDS("Completeness_data/wrapped_gridCaribbean.rds")
wrapped_gridIndoPacific <- readRDS("Completeness_data/wrapped_gridIndoPacific.rds")

# Completeness with iNEXT and new grid ####

#Caribbean region
names(Caribbean_Cell_Assembly) #each row is a species (2254), and columns 2:10725 is the cell ID

listCaribbean<-as.data.frame(Caribbean_Cell_Assembly[,2:ncol(Caribbean_Cell_Assembly)])

#iNterpolation and EXTrapolation for the first three Hill numbers (see Chao et al. 2004):
Caribbean_iNEXT<-iNEXT(listCaribbean,q=c(0,1,2), 
                           datatype = "abundance", 
                           endpoint = NULL, #double the reference sample size
                           conf=0.95, nboot=500)
saveRDS(Caribbean_Completeness, "Completeness_output/Caribbean_Completeness.rds") #DONE

CompleteCaribbeanCells <- Caribbean_iNEXT$iNextEst$coverage_based %>%
  filter(Method == "Observed" & Order.q %in% 0 & SC >= 0.9)  

#Exclude cells in SS with less than 4 species
Caribbean_SS = Caribbean_SS[Caribbean_SS$cell %in% CompleteCaribbeanCells$Assemblage, ]

saveRDS(Caribbean_SS, "Completeness_output/Caribbean_SS_Completeness.rds") #DONE

#Get the grid cell boundaries for cells with completeness >0.9
wrapped_gridCaribbean = wrapped_gridCaribbean[wrapped_gridCaribbean$seqnum %in% CompleteCaribbeanCells$Assemblage, ]

gridCaribbean = merge(wrapped_gridCaribbean,CompleteCaribbeanCells,by.x="seqnum",by.y="Assemblage")

saveRDS(gridCaribbean, "Completeness_output/gridCaribbean.rds") #DONE

#Now with the IndoPacific
listIndoPacific<-as.data.frame(IndoPacific_Cell_Assembly[,2:ncol(IndoPacific_Cell_Assembly)])

#iNterpolation and EXTrapolation for the first three Hill numbers (see Chao et al. 2004):
IndoPacific_iNEXT<-iNEXT(listIndoPacific,q=c(0,1,2), 
                         datatype = "abundance", 
                         endpoint = NULL, #double the reference sample size
                         conf=0.95, nboot=500)
saveRDS(IndoPacific_iNEXT, "Completeness_output/IndoPacific_Completeness.rds") #DONE

CompleteIndoPacificCells <- IndoPacific_iNEXT$iNextEst$coverage_based %>%
  filter(Method == "Observed" & Order.q %in% 0 & SC >= 0.9)  

#Exclude cells in SS with less than 4 species
IndoPacific_SS = IndoPacific_SS[IndoPacific_SS$cell %in% CompleteIndoPacificCells$Assemblage, ]

saveRDS(IndoPacific_SS, "Completeness_output/IndoPacific_SS_Completeness.rds") #DONE

#Get the grid cell boundaries for cells with completeness >0.9
wrapped_gridIndoPacific = wrapped_gridIndoPacific[wrapped_gridIndoPacific$seqnum %in% CompleteIndoPacificCells$Assemblage, ]

gridIndoPacific = merge(wrapped_gridIndoPacific,CompleteIndoPacificCells,by.x="seqnum",by.y="Assemblage")
saveRDS(gridIndoPacific, "Completeness_output/gridIndoPacific.rds") #DONE

#End of this code
```

## Example: Bananaquit populations

We selected a widespread distributed species to test our predictions of
difference in stationary dynamic abundance from the mainland to islands,
the Bananaquit *Coereba flaveola*

```{r eval=FALSE}
library(tidyverse) #seven packages in one for data management and visualization
library(MASS); #R functions and datasets to support "Modern Applied Statistics with S", a book from W.N. Venables and B.D. Ripley
source("ROUSSE-2.0.R") #Code from Dennis & Ponciano Ecology 2014.

#Call data and manipulate it 

Coefla <- readRDS("Completeness_output/Caribbean_SS_Completeness_elevation.rds") |>
  ungroup() |>
  filter(scientific_name == "Coereba flaveola") |>
  mutate(Day.t = as.numeric(observation_date)-15339, #First estimation of estimate in a year basis
         Site.k = as.character(cell), 
         Year.t = as.numeric(year)) 

#Filter cells with more than 3 years of sampling
CellsMore3yrs <- Coefla %>% 
  group_by(Site.k, Year.t) %>%
  summarise(Lists = n()) %>% 
  group_by(Site.k) %>% 
  summarise(List_Year = n()) %>%
  filter(List_Year >= 3)

#Select only cells with more than 3 years of data
Coefla = Coefla[Coefla$Site.k %in% CellsMore3yrs$Site.k, ]

#Mean count per year per site
Coefla <- Coefla |>
  group_by(Site.k, Year.t) |>
  summarise(Observed.year.y = round(mean(observation_count),0)) |>
  left_join(Coefla, by = c("Site.k","Year.t"))

#Mean count per day per site

Coefla <- Coefla |>
  group_by(Site.k, Day.t) |>
  summarise(Observed.day.y = round(mean(observation_count),0)) |>
  left_join(Coefla, by = c("Site.k","Day.t"))

Coefla |> 
  dplyr::select(state_code, Site.k, Year.t, Day.t, Observed.year.y, Observed.day.y, observation_count) |>
  arrange(state_code) |>
  head()

saveRDS(Coefla, file = "Coereba_flaveola_data.rds")
```

We can use the map of coast line Mainlan-Island map from Sayer *et al.*
(J. Oper. Oceanogr.)

```{r warning=FALSE}
library(sf); #deal with shape files
library(tidyverse)
library(maps); #load maps
library(ggmagnify); # Zoom detail (inset figures)

Coefla <- readRDS("Coereba_flaveola_data.rds")

#The grid of cells
gridCaribbeanPost <- readRDS("Completeness_output/gridCaribbean.rds") #3823 Sites

#Only have the values in Coefla
gridCaribbeanPost = gridCaribbeanPost[gridCaribbeanPost$seqnum %in% Coefla$Site.k, ] #1033 Sites (cells k)

#A world map as reference

world1 <- sf::st_as_sf(map(database = 'world', plot = FALSE, fill = TRUE))
world1

#Global Shoreline Vector
#coast <- st_read("USGSEsriWCMC_GlobalIslands_v3/v10/globalislandsfix.gdb") 

#Different projections!
#Extract CRS of world1 map
world1_crs <- st_crs(world1)
#Transform the geodatabase to match the CRS projection of world1
#coastal <- st_transform(coast, world1_crs)

ggplot() +
  geom_sf(data = world1)+
  geom_sf(data=gridCaribbeanPost,
          aes(color = SpRichness, 
              fill = SpRichness)) +
  scale_color_gradient2(low = alpha("#4575b4", 0.95), 
                        mid = alpha("#ffffbf", 0.95), midpoint = 73,
                        high = alpha("#d73027", 0.95))+
  scale_fill_gradient2(low = alpha("#4575b4", 0.75), 
                       mid = alpha("#ffffbf", 0.75), midpoint = 73,
                       high = alpha("#d73027", 0.75))+
  geom_point(data = Coefla, aes(x = longitude, y = latitude), alpha = 0.5, size = 0.001)+
  coord_sf(xlim = c(-91, -59.4), 
           ylim =  c(8, 27)) +
  scale_x_continuous(breaks = seq(from = -90, to = -60, by = 15))+
  labs(y = "Latitude",
       x = "Longitude",       
       tag = "",
       title = "Neotropical region",
       subtitle = "Caribbean (high eBird coverage-based completeness)",
       color = "Species richness",
       fill = "Species richness") +
  theme_classic()+
  theme(legend.position = "bottom",
        legend.direction = "horizontal")+
  guides(fill = F)+
  geom_magnify(data = Coefla, aes(x = longitude, y = latitude), 
               from = c(xmin = -67.3, xmax = -65.6, ymin = 17.9, ymax = 18.6), 
               to = c(xmin = -70, xmax = -59, ymin = 22, ymax = 26.5),
               shadow = TRUE) 

```

So, how is the time-series of the observed process?

```{r warning=FALSE}

gridCaribbeanPost$Site.k <- as.character(gridCaribbeanPost$seqnum)

Coefla <- Coefla |>
  left_join(gridCaribbeanPost)

ggplot(Coefla, aes(x = observation_date, group = Site.k))+
  geom_line(aes(y = log(Observed.day.y), color = SpRichness), alpha = 0.25, linetype = "dotted")+
  geom_line(aes(y = log(Observed.year.y), color = SpRichness), alpha = 0.25, linetype = "dashed")+
  scale_color_gradient2(low = alpha("#4575b4", 0.95), 
                        mid = alpha("#ffffbf", 0.95), midpoint = 73,
                        high = alpha("#d73027", 0.95))+
  geom_point(aes(y = log(Observed.day.y)), alpha = 0.01, size = 0.05, shape = 1)+
  geom_point(aes(y = log(Observed.year.y)), alpha = 0.01, size = 0.1, shape = 2)+
  scale_y_continuous(expand = c(0,0))+
  scale_x_date(expand = c(0,0))+
  labs(x = "Observation date",
       y = "Log(mean observed)",
       color = "Species richness")+
  theme_classic()+
  theme(legend.position = "bottom")

```

This is very messy an difficult to see, lets reduce to values around the
Min., 1st Qu., Median, Mean, 3rd Qu., and Max. values of the Species
richness gradient.

```{r}

#Use some examples that represent sp richness:  different species richness

summary(Coefla$SpRichness)

CoeflaSimp <- Coefla |>
  filter(SpRichness %in% c(12,14,16:29, 31:68,72,
                           104,106,112,115,138,
                           155,156,160,174,179,186:188,196,
                           200:202,205:209,226:231,235,237,243:247,
                           250:253,261,270,278:284,293:299,
                           303:322,330:380)) |> #Select cells with variable richness and more than 200 observations
  group_by(SpRichness) |>
  sample_n(200) |>
  as.data.frame()


ggplot(CoeflaSimp, aes(x = observation_date, group = Site.k))+
  geom_line(aes(y = log(Observed.day.y), color = SpRichness), linetype = "dotted")+
  geom_line(aes(y = log(Observed.year.y), color = SpRichness), linetype = "dashed")+
  scale_color_gradient2(low = alpha("#4575b4", 0.95), 
                        mid = alpha("#ffffbf", 0.95), midpoint = 73,
                        high = alpha("#d73027", 0.95))+
  geom_point(aes(y = log(Observed.day.y)), size = 0.05, shape = 1)+
  geom_point(aes(y = log(Observed.year.y)), size = 0.1, shape = 2)+
  scale_y_continuous(expand = c(0,0))+
  scale_x_date(expand = c(0,0))+
  labs(x = "Observation date",
       y = "Log(mean observed)",
       color = "Species richness")+
  theme_classic()+
  theme(legend.position = "bottom")


```

In the previous figure, we assumed that the mean $\mu_{t,k}$ of counts
each year (or day in the future) $t$ in each site $k$ represents the stationary
distribution of the stochastic density dependence in the population.

Look at the lower values of observed abundance in high richness sites,
and higher estimates in low richness sites. Could we try to plug-in
these time-series in the Gompertz state-state model? The problem is that
this data is not uniform... so we need the diffusion process of
Ornstein-Uhlenbeck process (Dennis & Ponciano, *Ecology* 2018).

Lele *et al.* (*Ecol. Lett.* 2007) provide the example of the GSS model
as a stochastic, density-dependent model (based on Dennis *et al.*,
*Ecol. Monogr.* 2006). In this model, the latent variable of $N_t$ as
the true abundance at time $t$ in the form: $$
N_t = N_{t-1}* {\rm exp} \left(a + b * {\rm ln}N_{t-1} + E_t \right)
$$ where $a$ is the intrinsic growth rate (a constant) and $b$ is a
density dependence parameter (another constant), while the environmental
stochasticity is $E_t \sim {\rm Normal}(0,\sigma^2)$. The variance
$\sigma^2$ is the process noise or environmental variability of the
system (Lele *et al.*, *Ecol. Lett.* 2007), where higher values could
serve to identify catastrophic stochasticity (Dennis *et al.*, *Ecol.
Monogr.* 1991).

Thus, we can define $X_t = {\rm ln}(N_t)$, or the logarithm of the
unobserved population abundances at time $t$, $t=(0,1,...,q)$. Applying
the logarithm scale:

$$
X_t = X_{t-1} + a + (b*X_{t-1}) + E_t
$$ factorizing out, $$
X_t = a + X_{t-1}(1+b) + E_t
$$ and simplify the terms, $c=1+b$ defines the density dependence
parameter:

$$
X_t = a + cX_{t-1} + E_t
$$

In the GSS model, the probability distribution of $X_t$ is normal with
mean and variance that change in function of time. We begin with a first
value $X_0$, that is plug-in in $X_1 = a + c(X_0) +E_1$, then this $X_1$
is plug-in in $X_2 = a + c(X_1) + E_2 = a + c(a + c(X_0) +E_1) + E_2$,
and so on... eventually, when $X_t \rightarrow{\infty}$, $X_t$
approaches a time-independent stationary distribution (carrying
capacity):

$$
X_{\infty} \sim {\rm Normal} \biggl(\mu=\frac{a}{(1-c)}, {\rm var} = \frac{\sigma^2}{(1-c)^2} \biggr)
$$ Thus, we can assume that our $X_0$ is part of that stationary
distribution for the long-run time-series.

The observations $Y_t$ are related to the true population abundances
$X_t$ by: $Y_t = X_t + F_t$, where $F_t \sim {\rm Normal}(0,\tau^2)$.
Here, the variance $\tau^2$ quantifies the amount of measurement error
or observation error. We have to compute the MLE for
$\Theta = [a, c, \sigma, \tau]$. However, we can also plug the
observations $Y_t$ as a Poisson process ($Y_t \sim {\rm Poisson}(N_t)$,
or in log-scale $Y_t \sim {\rm Poisson}({\rm ln}(X_t))$), reducing a
parameter estimation ($\tau$). However, the nature of the data from
eBird require account for the variaton of observation error in the
state-space model. In addition, this data is unequal in time intervals,
requiring including the diffusion Ornstein-Uhlenbeck process (Dennis &
Ponciano, *Ecology* 2018).

```{r}
head(CoeflaSimp)

CoeflaSimpAnual <- CoeflaSimp |>
  dplyr::select(Site.k, Year.t, Observed.year.y, SpRichness, country) |>
  unique() |>
  filter(Site.k != "2737266")

table(CoeflaSimpAnual$Site.k, CoeflaSimpAnual$Year.t) #this shows that some cells have only 1 or less observations

##Filter cells with more than 6 years of sampling
CellsMore6yrs <- CoeflaSimpAnual %>% 
  group_by(Site.k, Year.t) %>%
  summarise(Lists = n()) %>% 
  group_by(Site.k) %>% 
  summarise(List_Year = n()) %>%
  filter(List_Year >= 6)


#Select only cells with more than 6 years of data (from 3736 to 2695)
CoeflaSimpAnual = CoeflaSimpAnual[CoeflaSimpAnual$Site.k %in% CellsMore6yrs$Site.k, ]

#Chronological order
CoeflaSimpAnual <- CoeflaSimpAnual[order(CoeflaSimpAnual$Site.k, CoeflaSimpAnual$Year.t),]



ggplot(CoeflaSimpAnual, aes(x = Year.t, group = Site.k))+
  geom_line(aes(y = log(Observed.year.y), color = SpRichness), linetype = "dashed")+
  scale_color_gradient2(low = alpha("#4575b4", 0.95), 
                        mid = alpha("#ffffbf", 0.95), midpoint = 73,
                        high = alpha("#d73027", 0.95))+
  geom_point(aes(y = log(Observed.year.y)), size = 0.1, shape = 2)+
  scale_y_continuous(expand = c(0,0))+
  scale_x_continuous(expand = c(0,0), breaks = c(2012, 2014, 2016, 2018, 2020, 2022))+
  labs(x = "Observation date",
       y = "Log(mean observed)",
       color = "Species richness")+
  theme_classic()+
  theme(legend.position = "bottom")

```
onlye some sites to represent the process and run the model


```{r eval=FALSE}
PopID<- unique(CoeflaSimpAnual$Site.k) 
#308 cells for conducting the example

#empty list per PopID
Coefla_Parm <- data.frame(cell = character(), 
                          mu.hat = numeric(),
                          theta.hat = numeric(),
                          beta.hat = numeric(),
                          tau.hat = numeric())

for(PopID in unique(CoeflaSimpAnual$Site.k)){
  
  dt_tmp <- CoeflaSimpAnual[CoeflaSimpAnual$Site.k == PopID, ]
  
  Observed.t <- dt_tmp$Observed.year.y
  Time.t <- dt_tmp$Year.t
  #-------- Log-transform the observations to carry all the calculations in this program--------------------#
  log.obs    <- log(Observed.t); 
  #---------------------------------------------------------------------------------------------------------#
  
  #----------------------------------------------------------------------
  #        PARAMETER ESTIMATION, PARAMETRIC BOOTSTRAP AND PREDICTIONS
  #----------------------------------------------------------------------
  # Before doing the calculations, the user has to specify ONLY the following 4 options:
  
  # 1. Do you want to compute the ML estimates or the REML estimates?
  
  method <- "REML" # alternatively, set method <- "ML"
  
  # 2. Do you want to plot the predictions?
  pred.plot <- "FALSE" # Set it to "FALSE" if you do not want to plot the predictions
  
  # 3. Do you want to plot the parametric bootstrap distribution of the estimates?
  pboot.plot <- "FALSE" # Set it to "FALSE" if you do not want to plot the bootstrap distribution of the estimates
  
  # 4. How many bootstrap replicates?
  NBoot <- 10; # 10 just to try, for formal results use 1000 or more 
  
  #-------------------------------------------------------------------------------------------------#
  #  5. RUN THE FOLLOWING LINE OF CODE TO COMPUTE THE ESTIMATES, PREDICTIONS,
  #     AND RUN A PARAMETRIC BOOTSTRAP. THE USER DOES NOT NEED TO MODIFY THIS LINE OF CODE.
  #     THE OUTPUT OF THE FUNCTION 'ROUSS.CALCS' IS A LIST AND THE USER CAN RETRIEVE EACH OF THE   
  #     LIST ELEMENTS PRINTED AND SAVED IN THE OBJECT "all.results". 
  #     THE 95% PARAMETRIC BOOTSTRAP FOR BOTH, THE PARAMETERS AND THE PREDICTIONS ARE COMPUTED 
  #     BY THE FUNCTION "ROUSS.CALCS" 	
  #      
  #-------------------------------------------------------------------------------------------------#
  
  all.results <- ROUSS.CALCS(Yobs=log.obs,Tvec=Time.t, pmethod=method, nboot=NBoot, plot.pred=pred.plot, plot.bootdists = pboot.plot)
  
  Coefla_Parm <- rbind(Coefla_Parm, c(cell = PopID, all.results$parms.est))
  
  print(all.results$parms.est)
}

#It changes the column name, but save the estimates!!
colnames(Coefla_Parm) = c("cell", "mu.hat", "theta.hat", "beta.hat", "tau.hat")

saveRDS(Coefla_Parm, "Coefla_ParametersOUSS_abundance.rds")

```

$\hat{\mu}$ is the mean stationary log-abundance estimate for each site!

We can also predict the estimated abundance for each year:

```{r eval=FALSE}

PopID<- unique(CoeflaSimpAnual$Site.k) 

GSS.Ban <- list()

for(PopID in unique(CoeflaSimpAnual$Site.k)){
  
  ParmPop <- Coefla_Parm %>%
    filter(cell == PopID) %>%
    dplyr::select(mu.hat, theta.hat, beta.hat, tau.hat)

    dt_tmp <- CoeflaSimpAnual[CoeflaSimpAnual$Site.k == PopID, ]
  
    Observed.t <- dt_tmp$Observed.year.y
    Tvec <- dt_tmp$Year.t
  #-------- Log-transform the observations to carry all the calculations in this program--------------------#
    Yobs <- log(Observed.t); 

    ROUSS.predict(parms = as.numeric(ParmPop), Yobs = Yobs, Tvec = Tvec, plot.it = "TRUE")

  PRBan.parms <- c(as.numeric(ParmPop))
  #Parameteric Bootstrap function and confidence intervals
  B = NBoot

  tt <- Tvec-Tvec[1]
  long.t <- tt[1]:max(tt)
  nparms    <- length(PRBan.parms);
  preds.boot1<- matrix(0,nrow=B,ncol=length(tt))
  preds.boot2<- matrix(0,nrow=B,ncol=length(long.t))

  boot.remles <- matrix(0,nrow=B,ncol=nparms+1); 
  all.sims  <- ROUSS.sim(nsims=B,parms=PRBan.parms,Tvec=Tvec);
  all.preds <- ROUSS.predict(parms=PRBan.parms, Yobs=Yobs,Tvec=Tvec,plot.it="FALSE")
  reml.preds<- all.preds[[1]][,2]
  reml.longpreds <- all.preds[[2]][,2]		

  for(b in 1:B ){
  
    bth.timeseries <- all.sims[,b];
    remles.out <- ROUSS.REML(Yobs=bth.timeseries,Tvec=Tvec, parms.guess=PRBan.parms);
    boot.remles[b,] <- c(remles.out$remls, remles.out$lnLhat);
    all.bootpreds <- ROUSS.predict(parms=remles.out$remls, Yobs=bth.timeseries,Tvec=Tvec,plot.it="FALSE");
    preds.boot1[b,] <- all.bootpreds[[1]][,2]
    preds.boot2[b,] <- all.bootpreds[[2]][,2]
  
  } 

CIs.mat <- apply(boot.remles,2,FUN=function(x){quantile(x,probs=c(0.025,0.975))});
CIs.mat <- rbind(CIs.mat[1,1:4],PRBan.parms,CIs.mat[2,1:4]);
rownames(CIs.mat) <- c("2.5%","REMLE","97.5%");
colnames(CIs.mat) <- c("mu", "theta","betasq","tausq");

preds.CIs1 <- apply(preds.boot1,2,FUN=function(x){quantile(x,probs=c(0.025,0.975))});
mean.boots <- apply(preds.boot1,2,FUN=function(x){quantile(x,probs=0.50)})
preds.CIs1 <- t(rbind(Tvec,reml.preds-(mean.boots-preds.CIs1[1,]), reml.preds, reml.preds+(preds.CIs1[2,]-mean.boots)));
colnames(preds.CIs1) <- c("Year","2.5%","REMLE","97.5%");

preds.CIs2 <- apply(preds.boot2,2,FUN=function(x){quantile(x,probs=c(0.025,0.975))});
mean.boots2 <- apply(preds.boot2,2,FUN=function(x){quantile(x,probs=0.50)})
preds.CIs2 <- t(rbind(long.t+Tvec[1],reml.longpreds-(mean.boots2-preds.CIs2[1,]), reml.longpreds, reml.longpreds+(preds.CIs2[2,]-mean.boots2)));

#preds.CIs2 <- apply(preds.boot2,2,FUN=function(x){quantile(x,probs=c(0.025,0.5,0.975))});
#preds.CIs2 <- t(rbind(long.t+Tvec[1],preds.CIs2));
colnames(preds.CIs2) <- c("Year","2.5%","REMLE","97.5%");
#pred.CIs2 <- cbind(preds.CIs2[,1], reml.longpreds-(preds.CIs2[,3]-preds.CIs2[,2]),reml.longpreds,reml.longpreds+(preds.CIs2[,4]-preds.CIs2[,3]))

boot.list <- list(boot.remles = boot.remles, CIs.mat = CIs.mat, preds.CIs1 = preds.CIs1, 
                  preds.CIs2=preds.CIs2)


boot.list$preds.CIs1

df.boot <- as.data.frame(cbind(PopID, boot.list$preds.CIs1))

GSS.Ban <- c(GSS.Ban, list(df.boot))

}


#extract estimates
GSS.BanAll <- bind_rows(GSS.Ban)
colnames(GSS.BanAll) <-c("Site.k", "Year.t","Low","REMLE","High");

GSS.BanAll <- GSS.BanAll %>%
  mutate_at(c("Year.t","Low","REMLE","High"), as.numeric) %>%
  left_join(CoeflaSimpAnual, by = c("Site.k", "Year.t"))

saveRDS(GSS.BanAll, "Coefla_predictedAbundance_REMLE_OUSS.rds")
```


```{r}

GSS.BanAll <- readRDS("Coefla_predictedAbundance_REMLE_OUSS.rds")

ggplot(GSS.BanAll, aes(x = Year.t))+
  geom_line(aes(y = log(Observed.year.y), group = Site.k, color = SpRichness), linetype = "dotted")+
  geom_point(aes(y = log(Observed.year.y), fill = SpRichness), shape = 21)+
  geom_line(aes(y = log(REMLE), group = Site.k, color = SpRichness), linetype = "solid")+
  geom_point(aes(y = log(REMLE), fill = SpRichness), shape = 22)+
  scale_color_gradient2(low = alpha("#4575b4", 0.95), 
                        mid = alpha("#ffffbf", 0.95), midpoint = 73,
                        high = alpha("#d73027", 0.95))+
  scale_fill_gradient2(low = alpha("#4575b4", 0.95), 
                        mid = alpha("#ffffbf", 0.95), midpoint = 73,
                        high = alpha("#d73027", 0.95))+
  labs(x = "t",
       y = "log(Observed / estimated abundance)",
       color = "Species richness",
       fill = "Species richness",
       title = "62 sites (hexagonal cells)")+
  theme_classic()+
  theme(legend.position = c(0.8,0.8),
        legend.direction = "horizontal")

ggplot(GSS.BanAll, aes(x = SpRichness, fill = SpRichness))+
  geom_point(aes(y = REMLE), shape = 22)+
  geom_point(aes(y = Observed.year.y), shape = 21)+
  geom_smooth(aes(x = SpRichness, y = REMLE), method = "lm", linetype = "solid")+
  geom_smooth(aes(x = SpRichness, y = Observed.year.y), method = "lm", linetype = "dashed")+
  scale_fill_gradient2(low = alpha("#4575b4", 0.95), 
                        mid = alpha("#ffffbf", 0.95), midpoint = 73,
                        high = alpha("#d73027", 0.95))+
  labs(x = "Species richness",
       y = "Observed / estimated abundance",
       fill = "Species richness",
       title = "Relationship of abundance and species richness")+
  theme_classic()+
  theme(legend.position = c(0.75,0.75),
        legend.direction = "horizontal")

```

### We can map our data

```{r}

GSS.map.df <- GSS.BanAll %>%
  group_by(Site.k, Year.t, REMLE) %>%
  summarise(SpRichness = mean(SpRichness),
            lon = mean(longitude),
            lat = mean(latitude))

ggplot() +
  geom_sf(data = world1)+
  geom_point(data = GSS.map.df, 
             aes(x = lon, y = lat, size = REMLE, fill = SpRichness), alpha = 0.5, shape = 21)+
  scale_fill_gradient2(low = alpha("#4575b4", 0.75), 
                       mid = alpha("#ffffbf", 0.75), midpoint = 73,
                       high = alpha("#d73027", 0.75))+
  coord_sf(xlim = c(-91, -59.4), 
           ylim =  c(8, 27)) +
  scale_x_continuous(breaks = seq(from = -90, to = -60, by = 15))+
  labs(y = "Latitude",
       x = "Longitude",       
       tag = "",
       title = "Caribbean (high eBird coverage-based completeness)",
       subtitle = "Abundance estimation using GSS(OUSS) models for Bananaquit",
       fill = "Species richness",
       size = "Stationary abundance estimation") +
  theme_classic()+
  theme(legend.position = "bottom",
        legend.direction = "horizontal")+
  geom_magnify(data = GSS.map.df, aes(x = lon, y = lat), 
               from = c(xmin = -86.2, xmax = -81.2, ymin = 8, ymax = 13.5), 
               to = c(xmin = -75, xmax = -62, ymin = 10, ymax = 25),
               shadow = TRUE) 

```

End of this document.
Orlando.

## A previous approach (reducing the observer error parameter and applying Data Cloning)

We tried with the Poisson process (adjust to the length of the time
series). Here is the GSS.dc model:

```{r eval=FALSE}
StochGSS2.dc <- function(){

  # Priors on model parameters. Priors are DC1 in Lele et al (2007)
    a1 ~ dnorm(0,1);   # constant, the population growth rate. 
    c1 ~ dunif(-1,1);      # constant, the density dependence parameter. 
    sig1 ~ dlnorm(-0.5,10); #variance parameter of stochastic  environment (process noise) in the system
    stovar1 <- 1/pow(sig1,2)

    for(k in 1:K){
      # Simulate trajectory that depends on the previous
      mean_X1[1,k] <- a1/(1-c1) # Expected value of the first realization of the process
                          # this is drawn from the stationary distribution of the process
                          # Equation 14 (main text) and  A.4 in Appendix of Dennis et al 2006
      Varno1[k] <- pow(sig1,2)/(1-pow(c1,2)) #. Equation A.5 in Appendix of Dennis et al 2006
      
 
      # Updating the state: Stochastic process for 29 steps
      X1[1,k]~dnorm(mean_X1[1,k], 1/Varno1[k]); #first estimation of population
      
      #iteration of the GSS model in the data
      for (i in 2:12) {
        mean_X1[i,k] <- a1 + c1 * X1[(i - 1),k]
        X1[i,k] ~ dnorm(mean_X1[i,k], stovar1) # Et is included here since a+cX(t-1) + Et ~ Normal(a+cX(t-1),sigma^2)
      }
  
    # Updating the observations, from the counts
      for (i in 1:12) {
        Y1[i,k] ~ dpois(exp(X1[i,k])) #Here we can plug the probability of detection, p*exp(X1[i,k]), and we could also estimate it from the data!
      }
  }

}

```

Now, we select the data, first Costa Rica - Alejuela (CR-A), which has
complete data 2012-2023.

```{r eval=FALSE}
table(Pop$state_code, Pop$year)

#Costa Rica - Alejuela

CR.A <- Pop %>%
  dplyr::filter(state_code == "CR-A") %>%
  dplyr::select(mu_count)

CR <- round(CR.A$mu_count, 0)

datalistGSS.CR.dc <- list(K = 1,
                       Y1 = dcdim(data.matrix(CR))) # Not need to be in LOG!!!

dcrun.GSS2.CR <- dc.fit(data = datalistGSS.CR.dc,
                          params = c("a1", "c1", "sig1"), #extract from the model
                          model = StochGSS2.dc, 
                          n.clones = c(5,10,50,100,150),
                          multiply = "K",
                          n.chains = 10,
                          n.adapt = 10000,
                          n.update = 10000,
                          thin = 10,
                          n.iter = 100000)

summary(dcrun.GSS2.CR);

dcdiag(dcrun.GSS2.CR) 

```

Now, Aruba (Lesser Antilles), missing the two first years of data,
complete data 2014-2023

```{r eval=FALSE}

A <- Pop %>%
  dplyr::filter(state_code == "AW-") %>%
  dplyr::select(mu_count)

AW <- c(round(A$mu_count, 0)) #the

#and have to adjust the model
StochGSS2.AW.dc <- function(){

  # Priors on model parameters. Priors are DC1 in Lele et al (2007)
    a1 ~ dnorm(0,1);   # constant, the population growth rate. 
    c1 ~ dunif(-1,1);      # constant, the density dependence parameter. 
    sig1 ~ dlnorm(-0.5,10); #variance parameter of stochastic  environment (process noise) in the system
    stovar1 <- 1/pow(sig1,2)

    for(k in 1:K){
      # Simulate trajectory that depends on the previous
      mean_X1[1,k] <- a1/(1-c1) # Expected value of the first realization of the process
                          # this is drawn from the stationary distribution of the process
                          # Equation 14 (main text) and  A.4 in Appendix of Dennis et al 2006
      Varno1[k] <- pow(sig1,2)/(1-pow(c1,2)) #. Equation A.5 in Appendix of Dennis et al 2006
      
 
      # Updating the state: Stochastic process for 29 steps
      X1[1,k]~dnorm(mean_X1[1,k], 1/Varno1[k]); #first estimation of population
      
      #iteration of the GSS model in the data
      for (i in 2:10) {
        mean_X1[i,k] <- a1 + c1 * X1[(i - 1),k]
        X1[i,k] ~ dnorm(mean_X1[i,k], stovar1) # Et is included here since a+cX(t-1) + Et ~ Normal(a+cX(t-1),sigma^2)
      }
  
    # Updating the observations, from the counts
      for (i in 1:10) {
        Y1[i,k] ~ dpois(exp(X1[i,k])) #Here we can plug the probability of detection, p*exp(X1[i,k]), and we could also estimate it from the data!
      }
  }

}


datalistGSS.AW.dc <- list(K = 1,
                       Y1 = dcdim(data.matrix(AW))) # Not need to be in LOG!!!

dcrun.GSS2.AW <- dc.fit(data = datalistGSS.AW.dc,
                          params = c("a1", "c1", "sig1"), #extract from the model
                          model = StochGSS2.AW.dc, 
                          n.clones = c(5,10,50,100,150),
                          multiply = "K",
                          n.chains = 10,
                          n.adapt = 10000,
                          n.update = 10000,
                          thin = 10,
                          n.iter = 100000)

summary(dcrun.GSS2.AW);

dcdiag(dcrun.GSS2.AW) 

```

Let's try Puerto Rico - San Juan, which has *NA* in 2013 and 2014. I
will run from 2015 to 2023

```{r eval=FALSE}

PR.SJ <- Pop %>%
  dplyr::filter(state_code == "PR-SJ") %>%
  dplyr::select(mu_count)

PR <- c(round(PR.SJ$mu_count, 0)) #the

#and have to adjust the model
StochGSS2.PR.dc <- function(){

  # Priors on model parameters. Priors are DC1 in Lele et al (2007)
    a1 ~ dnorm(0,1);   # constant, the population growth rate. 
    c1 ~ dunif(-1,1);      # constant, the density dependence parameter. 
    sig1 ~ dlnorm(-0.5,10); #variance parameter of stochastic  environment (process noise) in the system
    stovar1 <- 1/pow(sig1,2)

    for(k in 1:K){
      # Simulate trajectory that depends on the previous
      mean_X1[1,k] <- a1/(1-c1) # Expected value of the first realization of the process
                          # this is drawn from the stationary distribution of the process
                          # Equation 14 (main text) and  A.4 in Appendix of Dennis et al 2006
      Varno1[k] <- pow(sig1,2)/(1-pow(c1,2)) #. Equation A.5 in Appendix of Dennis et al 2006
      
 
      # Updating the state: Stochastic process for 29 steps
      X1[1,k]~dnorm(mean_X1[1,k], 1/Varno1[k]); #first estimation of population
      
      #iteration of the GSS model in the data
      for (i in 2:10) {
        mean_X1[i,k] <- a1 + c1 * X1[(i - 1),k]
        X1[i,k] ~ dnorm(mean_X1[i,k], stovar1) # Et is included here since a+cX(t-1) + Et ~ Normal(a+cX(t-1),sigma^2)
      }
  
    # Updating the observations, from the counts
      for (i in 1:10) {
        Y1[i,k] ~ dpois(exp(X1[i,k])) #Here we can plug the probability of detection, p*exp(X1[i,k]), and we could also estimate it from the data!
      }
  }

}


datalistGSS.PR.dc <- list(K = 1,
                       Y1 = dcdim(data.matrix(PR))) # Not need to be in LOG!!!

dcrun.GSS2.PR <- dc.fit(data = datalistGSS.PR.dc,
                          params = c("a1", "c1", "sig1"), #extract from the model
                          model = StochGSS2.PR.dc, 
                          n.clones = c(5,10,50,100,150),
                          multiply = "K",
                          n.chains = 10,
                          n.adapt = 10000,
                          n.update = 10000,
                          thin = 10,
                          n.iter = 100000)

summary(dcrun.GSS2.PR);

dcdiag(dcrun.GSS2.PR) 

```

Let's try Venezuela - Miranda, which has data from 2016 to 2023

```{r eval=FALSE}

VE.M <- Pop %>%
  dplyr::filter(state_code == "VE-M") %>%
  dplyr::select(mu_count)

VE <- c(round(VE.M$mu_count, 0)) #the

#and have to adjust the model
StochGSS2.VE.dc <- function(){

  # Priors on model parameters. Priors are DC1 in Lele et al (2007)
    a1 ~ dnorm(0,1);   # constant, the population growth rate. 
    c1 ~ dunif(-1,1);      # constant, the density dependence parameter. 
    sig1 ~ dlnorm(-0.5,10); #variance parameter of stochastic  environment (process noise) in the system
    stovar1 <- 1/pow(sig1,2)

    for(k in 1:K){
      # Simulate trajectory that depends on the previous
      mean_X1[1,k] <- a1/(1-c1) # Expected value of the first realization of the process
                          # this is drawn from the stationary distribution of the process
                          # Equation 14 (main text) and  A.4 in Appendix of Dennis et al 2006
      Varno1[k] <- pow(sig1,2)/(1-pow(c1,2)) #. Equation A.5 in Appendix of Dennis et al 2006
      
 
      # Updating the state: Stochastic process for 29 steps
      X1[1,k]~dnorm(mean_X1[1,k], 1/Varno1[k]); #first estimation of population
      
      #iteration of the GSS model in the data
      for (i in 2:9) {
        mean_X1[i,k] <- a1 + c1 * X1[(i - 1),k]
        X1[i,k] ~ dnorm(mean_X1[i,k], stovar1) # Et is included here since a+cX(t-1) + Et ~ Normal(a+cX(t-1),sigma^2)
      }
  
    # Updating the observations, from the counts
      for (i in 1:9) {
        Y1[i,k] ~ dpois(exp(X1[i,k])) #Here we can plug the probability of detection, p*exp(X1[i,k]), and we could also estimate it from the data!
      }
  }

}


datalistGSS.VE.dc <- list(K = 1,
                       Y1 = dcdim(data.matrix(VE))) # Not need to be in LOG!!!

dcrun.GSS2.VE <- dc.fit(data = datalistGSS.VE.dc,
                          params = c("a1", "c1", "sig1"), #extract from the model
                          model = StochGSS2.VE.dc, 
                          n.clones = c(5,10,50,100,150),
                          multiply = "K",
                          n.chains = 10,
                          n.adapt = 10000,
                          n.update = 10000,
                          thin = 10,
                          n.iter = 100000)

summary(dcrun.GSS2.VE);
dcdiag(dcrun.GSS2.VE)

```

Let's try Grenada - Saint Andrew, which has six years of data

```{r eval=FALSE}

GR.SA <- Pop %>%
  dplyr::filter(state_code == "GD-01") %>%
  dplyr::select(mu_count)

GR <- c(round(GR.SA$mu_count, 0)) #the

#and have to adjust the model
StochGSS2.GR.dc <- function(){

  # Priors on model parameters. Priors are DC1 in Lele et al (2007)
    a1 ~ dnorm(0,1);   # constant, the population growth rate. 
    c1 ~ dunif(-1,1);      # constant, the density dependence parameter. 
    sig1 ~ dlnorm(-0.5,10); #variance parameter of stochastic  environment (process noise) in the system
    stovar1 <- 1/pow(sig1,2)

    for(k in 1:K){
      # Simulate trajectory that depends on the previous
      mean_X1[1,k] <- a1/(1-c1) # Expected value of the first realization of the process
                          # this is drawn from the stationary distribution of the process
                          # Equation 14 (main text) and  A.4 in Appendix of Dennis et al 2006
      Varno1[k] <- pow(sig1,2)/(1-pow(c1,2)) #. Equation A.5 in Appendix of Dennis et al 2006
      
 
      # Updating the state: Stochastic process for 29 steps
      X1[1,k]~dnorm(mean_X1[1,k], 1/Varno1[k]); #first estimation of population
      
      #iteration of the GSS model in the data
      for (i in 2:6) {
        mean_X1[i,k] <- a1 + c1 * X1[(i - 1),k]
        X1[i,k] ~ dnorm(mean_X1[i,k], stovar1) # Et is included here since a+cX(t-1) + Et ~ Normal(a+cX(t-1),sigma^2)
      }
  
    # Updating the observations, from the counts
      for (i in 1:6) {
        Y1[i,k] ~ dpois(exp(X1[i,k])) #Here we can plug the probability of detection, p*exp(X1[i,k]), and we could also estimate it from the data!
      }
  }

}


datalistGSS.GR.dc <- list(K = 1,
                       Y1 = dcdim(data.matrix(GR))) # Not need to be in LOG!!!

dcrun.GSS2.GR <- dc.fit(data = datalistGSS.GR.dc,
                          params = c("a1", "c1", "sig1"), #extract from the model
                          model = StochGSS2.GR.dc, 
                          n.clones = c(5,10,50,100,150),
                          multiply = "K",
                          n.chains = 10,
                          n.adapt = 10000,
                          n.update = 10000,
                          thin = 10,
                          n.iter = 100000)

summary(dcrun.GSS2.GR);

dcdiag(dcrun.GSS2.GR)

```

And Jamaica - Saint Andrew, which has four years of data

```{r eval=FALSE}

JM.SA <- Pop %>%
  dplyr::filter(state_code == "JM-02") %>%
  dplyr::select(mu_count)

JM <- c(round(JM.SA$mu_count, 0)) #the

#and have to adjust the model
StochGSS2.JM.dc <- function(){

  # Priors on model parameters. Priors are DC1 in Lele et al (2007)
    a1 ~ dnorm(0,1);   # constant, the population growth rate. 
    c1 ~ dunif(-1,1);      # constant, the density dependence parameter. 
    sig1 ~ dlnorm(-0.5,10); #variance parameter of stochastic  environment (process noise) in the system
    stovar1 <- 1/pow(sig1,2)

    for(k in 1:K){
      # Simulate trajectory that depends on the previous
      mean_X1[1,k] <- a1/(1-c1) # Expected value of the first realization of the process
                          # this is drawn from the stationary distribution of the process
                          # Equation 14 (main text) and  A.4 in Appendix of Dennis et al 2006
      Varno1[k] <- pow(sig1,2)/(1-pow(c1,2)) #. Equation A.5 in Appendix of Dennis et al 2006
      
 
      # Updating the state: Stochastic process for 29 steps
      X1[1,k]~dnorm(mean_X1[1,k], 1/Varno1[k]); #first estimation of population
      
      #iteration of the GSS model in the data
      for (i in 2:4) {
        mean_X1[i,k] <- a1 + c1 * X1[(i - 1),k]
        X1[i,k] ~ dnorm(mean_X1[i,k], stovar1) # Et is included here since a+cX(t-1) + Et ~ Normal(a+cX(t-1),sigma^2)
      }
  
    # Updating the observations, from the counts
      for (i in 1:4) {
        Y1[i,k] ~ dpois(exp(X1[i,k])) #Here we can plug the probability of detection, p*exp(X1[i,k]), and we could also estimate it from the data!
      }
  }

}


datalistGSS.JM.dc <- list(K = 1,
                       Y1 = dcdim(data.matrix(JM))) # Not need to be in LOG!!!

dcrun.GSS2.JM <- dc.fit(data = datalistGSS.JM.dc,
                          params = c("a1", "c1", "sig1"), #extract from the model
                          model = StochGSS2.JM.dc, 
                          n.clones = c(5,10,50,100,150),
                          multiply = "K",
                          n.chains = 10,
                          n.adapt = 10000,
                          n.update = 10000,
                          thin = 10,
                          n.iter = 100000)

summary(dcrun.GSS2.JM);

dcdiag(dcrun.GSS2.JM)

```

Now the null:

```{r eval=FALSE}


H0 <- round(Pop$mu_count, 0)

#and have to adjust the model
StochGSS2.H0.dc <- function(){

  # Priors on model parameters. Priors are DC1 in Lele et al (2007)
    a1 ~ dnorm(0,1);   # constant, the population growth rate. 
    c1 ~ dunif(-1,1);      # constant, the density dependence parameter. 
    sig1 ~ dlnorm(-0.5,10); #variance parameter of stochastic  environment (process noise) in the system
    stovar1 <- 1/pow(sig1,2)

    for(k in 1:K){
      # Simulate trajectory that depends on the previous
      mean_X1[1,k] <- a1/(1-c1) # Expected value of the first realization of the process
                          # this is drawn from the stationary distribution of the process
                          # Equation 14 (main text) and  A.4 in Appendix of Dennis et al 2006
      Varno1[k] <- pow(sig1,2)/(1-pow(c1,2)) #. Equation A.5 in Appendix of Dennis et al 2006
      
 
      # Updating the state: Stochastic process for 29 steps
      X1[1,k]~dnorm(mean_X1[1,k], 1/Varno1[k]); #first estimation of population
      
      #iteration of the GSS model in the data
      for (i in 2:51) {
        mean_X1[i,k] <- a1 + c1 * X1[(i - 1),k]
        X1[i,k] ~ dnorm(mean_X1[i,k], stovar1) # Et is included here since a+cX(t-1) + Et ~ Normal(a+cX(t-1),sigma^2)
      }
  
    # Updating the observations, from the counts
      for (i in 1:51) {
        Y1[i,k] ~ dpois(exp(X1[i,k])) #Here we can plug the probability of detection, p*exp(X1[i,k]), and we could also estimate it from the data!
      }
  }

}


datalistGSS.H0.dc <- list(K = 1,
                       Y1 = dcdim(data.matrix(H0))) # Not need to be in LOG!!!

dcrun.GSS2.H0 <- dc.fit(data = datalistGSS.H0.dc,
                          params = c("a1", "c1", "sig1"), #extract from the model
                          model = StochGSS2.H0.dc, 
                          n.clones = c(5,10,50,100,150),
                          multiply = "K",
                          n.chains = 10,
                          n.adapt = 10000,
                          n.update = 10000,
                          thin = 10,
                          n.iter = 100000)

summary(dcrun.GSS2.H0);

dcdiag(dcrun.GSS2.H0)

```

Let's combine the results

```{r eval=FALSE}

MLE <- as.data.frame(rbind(coef(dcrun.GSS2.H0),
             coef(dcrun.GSS2.CR),
             coef(dcrun.GSS2.VE),
             coef(dcrun.GSS2.AW),
             coef(dcrun.GSS2.GR),
             coef(dcrun.GSS2.PR),
             coef(dcrun.GSS2.JM)))

MLE$Pop <- c("Null",
             "CR-A",
             "VE-M",
             "AW-",
             "GD-01",
             "PR-SJ",
             "JM-02")

MLE$N_equ <- exp(-(MLE$a1/(MLE$c1-1)))

print(MLE)

#Some figures

#how many records per site?
Pop %>% 
  group_by(state_code) %>% 
  summarise(AllRecords = sum(n_records))
```

Mainland sites

```{r eval=FALSE}

par(mfrow = c(1,2));#create a 2x2 plotting matrix

hist(rnorm(10000, mean = MLE[2,5], sd = MLE[2,3]/(1-MLE[2,2])),
     xlab = "Equilibrium density",
     main = "Histogram of Costa Rica - Alejuela",
     xlim = c(0,10));
abline(v=MLE[2,5], col="blue");
abline(v=mean(CR), col = "red");
legend(5, 1500, legend=c("Abundance (estimated)", "Observation mean"),
       col=c("blue", "red"), lty=1, cex=0.8);
text(x = 8, y = 500, "n = 249")

hist(rnorm(10000, mean = MLE[3,5], sd = MLE[3,3]/(1-MLE[3,2])),
     xlab = "Equilibrium density",
     main = "Histogram of Venezuela - Miranda",
     xlim = c(0,10));
abline(v=MLE[3,5], col="blue");
abline(v=mean(VE), col = "red");
text(x = 8, y = 500, "n = 521")
```

Lesser Antilles sites

```{r eval=FALSE}

par(mfrow = c(1,2));#create a 2x2 plotting matrix

hist(rnorm(10000, mean = MLE[4,5], sd = MLE[4,3]/(1-MLE[4,2])),
     xlab = "Equilibrium density",
     main = "Histogram of Aruba",
     xlim = c(0,10));
abline(v=MLE[4,5], col="blue");
abline(v=mean(AW), col = "red");
text(x = 8, y = 500, "n = 229")

hist(rnorm(10000, mean = MLE[5,5], sd = MLE[5,3]/(1-MLE[5,2])),
     xlab = "Equilibrium density",
     main = "Histogram of Grenada - Saint Andrew",
     xlim = c(0,10));
abline(v=MLE[5,5], col="blue");
abline(v=mean(GR), col = "red");
text(x = 8, y = 500, "n = 117")
```

Greater Antilles sites

```{r eval=FALSE}
par(mfrow = c(1,2));#create a 2x2 plotting matrix

hist(rnorm(10000, mean = MLE[6,5], sd = MLE[6,3]/(1-MLE[6,2])),
     xlab = "Equilibrium density",
     main = "Histogram of Puerto Rico - San Juan",
     xlim = c(0,10));
abline(v=MLE[6,5], col="blue");
abline(v=mean(PR), col = "red");
text(x = 8, y = 500, "n = 105")

hist(rnorm(10000, mean = MLE[7,5], sd = MLE[7,3]/(1-MLE[7,2])),
     xlab = "Equilibrium density",
     main = "Histogram of Jamaica - Saint Andrew",
     xlim = c(0,10));
abline(v=MLE[7,5], col="blue");
abline(v=mean(JM), col = "red");
text(x = 8, y = 500, "n = 18")
```

Null hypothesis and relationship

```{r eval=FALSE}
par(mfrow = c(1,2));#create a 2x2 plotting matrix

hist(rnorm(10000, mean = MLE[1,5], sd = MLE[1,3]/(1-MLE[1,2])),
     xlab = "Equilibrium density",
     main = "Histogram of all the Meta-archipelago of the Caribbean",
     xlim = c(0,10));
abline(v=MLE[1,5], col="blue");
abline(v=mean(H0), col = "red");
text(x = 8, y = 500, "n = 1,239")

MLE$S <- c(NA, 149, 76, 67, 51, 39, 31)

plot(x = MLE[c(2:7),6], y = MLE[c(2:7),5], pch = 19, cex = 4,
     xlim = c(0,150), ylim = c(0,5),
     xlab = "Species richness",
     ylab = "Abundance (estimated)",
     main = "Hypothetical relationship",
     col = c("#F36F20", "#DD692E", "#864268", "#704C76", "#434174","#01458A"));
abline(lm(MLE$N_equ ~ MLE$S), col = "darkgray");
text(x = 25, y = 1, expression('R' ^ 2~"= 0.189; p = 0.215"))

```

### Likelihood Ratio test

The parameters for the estimation of the unknown, mean number of
Bananaquit per list of ≤2 km and 10-120 minutes in the six sites $s$
($s = CR, VE, AR, GD, PR, JM$) are $\mathbf{\Theta} = [a, c, \sigma]$.
Several possible scenarios regarding the comparison of these quantities
across sites comes to mind. The simple scenario is that all parameters
are equal among the meta-archipelago and representing identical
parameters $\mathbf{\Theta}$, thus our working null hypothesis is

$$
{\rm H_{0}} : \mathbf{\Theta}_{CR} = \mathbf{\Theta}_{VE} = \mathbf{\Theta}_{AW} = \mathbf{\Theta}_{GD} = \mathbf{\Theta}_{PR} = \mathbf{\Theta}_{JM} = \mathbf{\Theta}
$$

A very contrasting alternative hypothesis (Model1) represent differences
among the sites $s$ in the meta-archipelago, and is $$
{\rm H_{1}} : \mathbf{\Theta}_{CR} \neq \mathbf{\Theta}_{VE} \neq \mathbf{\Theta}_{AW} \neq \mathbf{\Theta}_{GD} \neq \mathbf{\Theta}_{PR} \neq \mathbf{\Theta}_{JM}
$$

A third hypothesis is that sub-regions within the meta-archipelago
(Mainland vs Lesser Antilles vs Greater Antilles) have similar
parameters $\mathbf{\Theta}_{Subregion}$, and they differ from the
others...

$$
{\rm H_{2}} : \mathbf{\Theta}_{CR} = \mathbf{\Theta}_{VE} \neq \mathbf{\Theta}_{AW} = \mathbf{\Theta}_{GD} \neq \mathbf{\Theta}_{PR} = \mathbf{\Theta}_{JM}
$$ or $$
{\rm H_{2}} : \mathbf{\Theta}_{Mainland} \neq \mathbf{\Theta}_{Lesser Antilles} \neq \mathbf{\Theta}_{Greater Antilles}
$$

And another hypothesis is that representing "the island syndrome"
(Mainland vs Islands) of difference in parameters
$\mathbf{\Theta}_{Mainland-Island}$:

$$
{\rm H_{3}} : \mathbf{\Theta}_{CR} = \mathbf{\Theta}_{VE} \neq \mathbf{\Theta}_{AW} = \mathbf{\Theta}_{GD} = \mathbf{\Theta}_{PR} = \mathbf{\Theta}_{JM}
$$ or $$
{\rm H_{3}} : \mathbf{\Theta}_{Mainland} \neq \mathbf{\Theta}_{Islands}
$$

The LR test statistics are $$
{\Lambda_{0-1}} = \frac{L_0(\mathbf{\hat\Theta})}{L_1(\mathbf{\hat\Theta}_{CR},\mathbf{\hat\Theta}_{VE},\mathbf{\hat\Theta}_{AW},\mathbf{\hat\Theta}_{GD},\mathbf{\hat\Theta}_{PR},\mathbf{\hat\Theta}_{JM})}
$$

$$
{\Lambda_{0-2}} = \frac{L_0(\hat\Theta)}{L_2(\mathbf{\hat\Theta}_{Mainland},\mathbf{\hat\Theta}_{Lesser Antilles},\mathbf{\hat\Theta}_{Greater Antilles})}
$$

$$
{\Lambda_{0-3}} = \frac{L_0(\hat\Theta)}{L_3(\mathbf{\hat\Theta}_{Mainland},\mathbf{\hat\Theta}_{Islands})}
$$

$$
{\Lambda_{2-1}} = \frac{L_2(\mathbf{\hat\Theta}_{Mainland},\mathbf{\hat\Theta}_{Lesser Antilles},\mathbf{\hat\Theta}_{Greater Antilles})}{L_1(\mathbf{\hat\Theta}_{CR},\mathbf{\hat\Theta}_{VE},\mathbf{\hat\Theta}_{AW},\mathbf{\hat\Theta}_{GD},\mathbf{\hat\Theta}_{PR},\mathbf{\hat\Theta}_{JM})}
$$
